---
title: "Add spatial resources to b3data data package"
author: "Ward Langeraert"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
# Load packages
library(tidyverse)     # Data wrangling and visualisation
library(frictionless)  # Create frictionless data package
library(sf)            # Spatial objects

# Source functions
source(here::here("source", "R", "add_manual_resource.R"))

# Data path and create directory if necessary
data_path <- here::here("data", "raw")
dir.create(data_path, showWarnings = FALSE, recursive = TRUE)

package_path <- here::here("data", "b3data_package")
dir.create(package_path, showWarnings = FALSE, recursive = TRUE)
```

# Goal

Add spatial data resources to b3data frictionless data package.

# Methods

This report focuses on adding spatial resources (e.g. GeoPackage, raster) to the **b3data** frictionless data package.

## Source Data

Spatial data need to be saved locally in the data package directory and then added manually to the descriptor file using `append()`.
We provide the reference grids for the occurrence cubes in the data package following this naming convention:

```
resource_name = "<grid><resolution>refgrid<region>"
```

Example:  
`mgrs10_refgrid_belgium` → MGRS 10 km reference grid Belgium

## Resource-level Metadata

Each dataset includes the following metadata (see: [Frictionless resource spec](https://docs.ropensci.org/frictionless/articles/data-resource.html#properties-implementation)):

- **format**: e.g. `gpkg`
- **title**: e.g. `"MGRS 10 km reference grid for Belgium"`
- **description**: concise explanation of content
- **sources**: source if applicable
- **licenses**: [CC0-1.0](https://creativecommons.org/publicdomain/zero/1.0/)

# Datasets
## MGRS 10 km reference grid Belgium

Load the data.

```{r}
# Read reference grid
utm10_bel <- st_read(file.path(data_path, "utm10_bel.shp"))

# Visualise
ggplot() + geom_sf(data = utm10_bel)
```

Clean data and transform cell codes to MGRS.
Convert to EPSG:32631 (WGS 84 / UTM zone 31N).

```{r}
mgrs10_refgrid_belgium <- utm10_bel %>%
  select(utm_tag = TAG, geometry) %>%
  mutate(mgrscode = ifelse(grepl("^[A-G]", utm_tag), paste0("31U", utm_tag),
                           paste0("32U", utm_tag))) %>%
  select(mgrscode, geometry) %>%
  st_transform(32631)
```

Write to data package and add metadata.

```{r}
st_write(mgrs10_refgrid_belgium,
         file.path(package_path, "mgrs10_refgrid_belgium.gpkg"),
         delete_dsn = TRUE)
```

```{r}
# Read package
b3data_package <- read_package(file.path(package_path, "datapackage.json"))

# Add resource to data package
mgrs10_resource <- list(
  name = "mgrs10_refgrid_belgium",
  path = "mgrs10_refgrid_belgium.gpkg",
  format = "gpkg",
  title = "MGRS 10 km reference grid Belgium",
  description = paste("MGRS 10 km reference grid for the mainland of Belgium.",
                      "EPSG:32631 (WGS 84 / UTM zone 31N)."),
  licenses = list(list(
    name = "CC0-1.0",
    path = "https://creativecommons.org/publicdomain/zero/1.0/",
    title = "Creative Commons Zero v1.0 Universal"
  ))
)

b3data_package <- add_manual_resource(
  b3data_package, mgrs10_resource, replace = TRUE
)

# Write package to directory
write_package(
  package = b3data_package,
  directory = package_path,
  compress = TRUE
)
```

## EEA 100 km reference grid Europe

> Coming soon
