---
title: "Create b3data data package for occurrence cube datasets"
author: "Ward Langeraert"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
# Load packages
library(tidyverse)     # Data wrangling and visualisation
library(frictionless)  # Create frictionless data package
library(rgbif)         # Create occurrence cubes from GBIF data
library(sf)            # Spatial objects

# Source functions
source(here::here("source", "R", "download_occ_cube.R"))

# Data path and create directory if necessary
data_path <- here::here("data", "raw")
dir.create(data_path, showWarnings = FALSE, recursive = TRUE)
```

# Goal

Create [frictionless](https://docs.ropensci.org/frictionless/) data package with example datasets (occurrence cubes) for the [b3verse](https://docs.b-cubed.eu/guides/b3verse/).

# Methods

This report focuses on creating the **b3data** frictionless data package for tabular datasets.  
A separate report adds spatial resources (e.g. GeoJSON, raster) to the same data package.

## Data Package Overview

- **Name**: `b3data`
- **Location**: `b3data-scripts/data/b3data_package/`
- **Published at**: [Zenodo](https://zenodo.org/) (DOI to be reserved)
- **Used in**: [b3verse](https://docs.b-cubed.eu/guides/b3verse/)
- **Importable in R via**: [`frictionless`](https://docs.ropensci.org/frictionless/) R package

## Source Data

Tabular data is downloaded from **GBIF** using the [`rgbif`](https://docs.ropensci.org/rgbif/) package.

Each dataset (resource) corresponds to an **occurrence cube** and follows this naming convention:

```
resource_name = "<taxon>cube<region>_<grid><resolution>"
```

Example:  
`bird_cube_belgium_mgrs10` → bird occurrences in Belgium using the MGRS grid (10 km resolution)

## Resource-level Metadata

Each dataset includes the following metadata (see: [Frictionless resource spec](https://docs.ropensci.org/frictionless/articles/data-resource.html#properties-implementation)):

- **title**: e.g. `"Occurrence cube for birds in Belgium (MGRS 10 km)"`
- **description**: concise explanation of content, region, time period, method
- **sources**: GBIF DOI
- **licenses**: [CC BY-NC 4.0](https://creativecommons.org/licenses/by-nc/4.0/) (based on GBIF source license)

## Package-level Metadata

Metadata for the entire data package (see: [Frictionless Data Package spec](https://specs.frictionlessdata.io/data-package/#metadata)):

- **name**: `b3data`
- **id**: reserved Zenodo DOI
- **title**: `b3data: Data resources for the b3verse`
- **description**: overview of purpose, content, and structure of the data package
- **keywords**: e.g. `"biodiversity", "data cube", "b3verse", "GBIF", "frictionless"`
- **licenses**: [CC BY-NC 4.0](https://creativecommons.org/licenses/by-nc/4.0/)
- **version**: semantic versioning string, e.g. 0.1.0
- **sources**: GitHub repository [b3data-scripts](https://github.com/b-cubed-eu/b3data-scripts)
- **contributors**: name, ORCID, and roles

# Datasets
## Birds in Belgium (MGRS 10 km)

As example dataset on small scale, we include an occurrence cube of birds in Belgium between 2000 en 2024.
The taxonomical resolution is "species" and the temporal resolution is "year".
For aggregation, we choose the MGRS grid at 10 km scale.

```{r}
# nolint start: line_length_linter.
bird_cube_query <- "SELECT
  \"year\",
  GBIF_MGRSCode(10000, decimalLatitude, decimalLongitude,
  COALESCE(coordinateUncertaintyInMeters, 1000)) AS mgrsCode,
  speciesKey,
  species,
  family,
  COUNT(*) AS n,
  MIN(COALESCE(coordinateUncertaintyInMeters, 1000)) AS minCoordinateUncertaintyInMeters,
  IF(ISNULL(family), NULL, SUM(COUNT(*)) OVER (PARTITION BY family)) AS familyCount
  FROM
  occurrence
  WHERE
  occurrenceStatus = 'PRESENT'
  AND NOT ARRAY_CONTAINS(issue, 'ZERO_COORDINATE')
  AND NOT ARRAY_CONTAINS(issue, 'COORDINATE_OUT_OF_RANGE')
  AND NOT ARRAY_CONTAINS(issue, 'COORDINATE_INVALID')
  AND NOT ARRAY_CONTAINS(issue, 'COUNTRY_COORDINATE_MISMATCH')
  AND countryCode = 'BE'
  AND \"year\" >= 2000
  AND \"year\" <= 2024
  AND speciesKey IS NOT NULL
  AND decimalLatitude IS NOT NULL
  AND decimalLongitude IS NOT NULL
  AND class = 'Aves'
  GROUP BY
  \"year\",
  mgrsCode,
  speciesKey,
  family,
  species
  ORDER BY
  \"year\" ASC,
  mgrsCode ASC,
  speciesKey ASC"
# nolint end

# Download and load cube
bird_cube_belgium_mgrs10_full <- download_occ_cube(
  sql_query = bird_cube_query,
  file = "bird_cube_belgium_mgrs10.csv",
  path = data_path,
  overwrite = FALSE
)
```

We only select the grid cells that belong to the MGRS 10 km reference grid for the mainland of Belgium (see b3data `mgrs10_refgrid_belgium.geojson`).

```{r}
# Read reference grid
utm10_bel <- st_read(file.path(data_path, "utm10_bel.shp"))

# Visualise
ggplot() + geom_sf(data = utm10_bel)
```

```{r}
# Only select grid cells from reference grid
bird_cube_belgium_mgrs10 <- bird_cube_belgium_mgrs10_full %>%
  filter(substring(mgrscode, 4) %in% utm10_bel$TAG)
```

We create the package and add the dataset.

```{r}
b3data_package <- create_package() %>%
  add_resource(
    resource_name = "bird_cube_belgium_mgrs10",
    data = bird_cube_belgium_mgrs10,
    title = "Occurrence cube for birds in Belgium (MGRS 10 km)",
    description = paste(
      "Occurrence cube for birds in Belgium between 2000 en 2024.",
      "The taxonomical resolution is 'species' and",
      "the temporal resolution is 'year'",
      "Spatial aggregation is done using the MGRS grid at 10 km scale.",
      "Only grid cells that fall within the 10 km MGRS reference grid for",
      "mainland Belgium (see b3data: `mgrs10_refgrid_belgium.geojson`) are",
      "included."),
    sources = c(
      title = "GBIF Occurrence Download",
      path = "https://doi.org/10.15468/dl.y3wpwk"
    ),
    licenses = c(
      name = " CC BY-NC 4.0",
      path = "https://creativecommons.org/licenses/by-nc/4.0/",
      title = "Creative Commons Attribution-NonCommercial 4.0 International"
    )
  )
```

## Mammals of Europe (EEA 100 km)

> Coming soon

# Write package

We add the package metadata.

```{r}
b3data_package <- append(b3data_package,
                         c(name = "b3data"),
                         after = 0)
b3data_package <- append(b3data_package,
                         c(id = "https://doi.org/10.5281/zenodo.15181097"),
                         after = 1)
b3data_package <- append(b3data_package,
                         c(title = "b3data: Data resources for the b3verse"),
                         after = 3)
b3data_package <- append(
  b3data_package,
  c(description = paste(
    "This data package contains data resources to be used across the b3verse",
    "(https://docs.b-cubed.eu/guides/b3verse/).",
    "This includes example datasets (occurrence cubes) as well as spatial",
    "resources like reference grids or raster data."
  )),
  after = 4)
b3data_package <- append(b3data_package,
                         c(keywords = c("data cubes", "b3verse", "frictionless",
                                        "biodiversity")),
                         after = 5)
b3data_package <- append(
  b3data_package,
  c(licenses = c(
    name = " CC BY-NC 4.0",
    path = "https://creativecommons.org/licenses/by-nc/4.0/",
    title = "Creative Commons Attribution-NonCommercial 4.0 International"
    )),
  after = 6)
b3data_package <- append(b3data_package,
                         c(version = "0.1.0"),
                         after = 7)
b3data_package <- append(
  b3data_package,
  c(sources = c(
    title = "b3data-scripts",
    path = "https://github.com/b-cubed-eu/b3data-scripts"
    )),
  after = 8)
b3data_package <- append(
  b3data_package,
  c(contributors = c(
    c(
      title = "Ward Langeraert",
      path = "https://orcid.org/0000-0002-5900-8109",
      email = "ward.langeraert@inbo.be",
      role = "author",
      organization = "Research Institute for Nature and Forest (INBO)"
      ),
    c(
      title = "Toon Van Daele",
      path = "https://orcid.org/0000-0002-1362-853X",
      role = "contributor",
      organization = "Research Institute for Nature and Forest (INBO)"
      )
    )),
  after = 9)

# Warning: append() drops the custom datapackage class.
# It can be added again by running b3data_package through create_package()
b3data_package <- create_package(b3data_package)
```

We write the package to the directory.

```{r}
# Create directory
package_path <- here::here("data", "b3data_package")
dir.create(package_path, showWarnings = FALSE, recursive = TRUE)

# Write package to directory
write_package(
  package = b3data_package,
  directory = package_path,
  compress = TRUE)
```
